Antigenic phenotypes and phylogenetic metrics improve forecasts of seasonal influenza A/H3N2 evolution

### 2020-08-04 manuscript/main.tex
Don't force large figures to have their own page
The eLife template handles placement of these figures without our intervention.
Reposition relative vaccine comparison as a figure supplement
Moves the relative vaccine comparison figure from the supplemental material to a
figure supplement, to make comparison between the absolute and relative
distances easier.

\begin{figure}[p]
  \begin{center}
  \includegraphics[width=\textwidth]{figures/validation_figure_simulated-simulated_sample_3_test_tree-normalized_fitness.pdf}
  \caption{
  Test of best model for simulated populations (true fitness) using 9 years previously unobserved test data and fixed model coefficients.
  A) The correlation of log estimated clade frequency fold change, $\log_{10}{\frac{\hat{x}(t + \Delta{t})}{x(t)}}$, and log observed clade frequency fold change, $\log_{10}{\frac{x(t + \Delta{t})}{x(t)}}$, shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
  B) The rank of the estimated best strain based on its distance to the future in the best model was in the top 20th percentile for 89\% of 18 timepoints, confirming that the model makes a good choice when forced to select a single representative strain for the future population.
  C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (gray shading) based on 100 bootstraps.
  D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
  The corresponding results for the naive model are shown in Supplemental Fig.~\ref{sup_fig:test_of_naive_model_for_simulated_populations}.
  }
  \label{fig:test_of_best_model_for_simulated_populations}
  \end{center}
\end{figure}

@@

\begin{figure}[p]
  \begin{center}
  \includegraphics[width=\textwidth]{figures/validation_figure_natural-natural_sample_1_with_90_vpm_sliding_test_tree-cTiter_x-ne_star.pdf}
  \caption{
  Test of best model for natural populations of H3N2 viruses, the composite model of HI antigenic novelty and mutational load.
  A) The correlation of estimated and observed clade frequency fold changes shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
  B) The rank of the estimated best strain based on its distance to the future for eight timepoints.
  The estimated best strain was in the top 20th percentile of observed closest strains for 100\% of timepoints.
  C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (gray shading) based on 100 bootstraps.
  D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
  The corresponding results for the naive model are shown in Supplemental Fig.~\ref{sup_fig:test_of_naive_model_for_natural_populations}.
  }
  \label{fig:test_of_best_model_for_natural_populations}
  \end{center}
\end{figure}

@@

\begin{figure}[htb]
  \begin{center}
  \includegraphics[width=\textwidth]{figures/vaccine-comparison.pdf}
  \caption{
    Observed distance to natural H3N2 populations one year into the future for each vaccine strain (green) and the observed (blue) and estimated closest strains to the future by the mutational load and LBI model (orange), the HI antigenic novelty and mutational load model (purple), and the naive model (black).
    Vaccine strains were assigned to the validation or test timepoint closest to the date they were selected by the WHO.
    The weighted distance to the future for each strain was calculated from their amino acid sequences and the frequencies and sequences of the corresponding population one year in the future.
    Vaccine strain names are abbreviated from A/Fujian/411/2002, A/Wellington/1/2004, A/California/7/2004, A/Wisconsin/67/2005, A/Brisbane/10/2007, A/Perth/16/2009, A/Victoria/361/2011, A/Texas/50/2012, A/Switzerland/9715293/2013, A/HongKong/4801/2014, A/Singapore/Infimh-16-0019/2016, and A/Switzerland/8060/2017.
  }
  \label{fig:vaccine_comparison}
  \end{center}
\end{figure}

We further evaluated our models' ability to estimate the closest strain to the next season's H3N2 population by comparing our best models' selections to the WHO's vaccine strain selection.
For each season when the WHO selected a new vaccine strain and one year of future data existed in our validation or test periods, we measured the observed distance of that strain's sequence to the future and the corresponding distances to the future for the observed closest strains.
We compared these distances to those of the closest strains to the future as estimated by our best models for the validation period (mutational load and LBI) and the test period (HI antigenic novelty and mutational load).
The mutational load and LBI model selected strains that were as close or closer to the future than the corresponding vaccine strain for 10 (83\%) of the 12 seasons with vaccine updates (Fig.~\ref{fig:vaccine_comparison}).
On average, the strains selected by this model were closer to future than the vaccine strain by 1.93 AAs (Supplemental Fig.~\ref{sup_fig:vaccine_comparison_relative_distance}).
For the two seasons that the model selected more distant strains than the vaccine strain, the mean distance relative to the vaccine strain was 1.58 AAs.
The HI antigenic novelty and mutational load model performed similarly by identifying strains as close or closer to the future for 11 (92\%) seasons with an average improvement over the vaccine strains of 2.33 AAs.
For the one season that the model selected a more distant strain, that selected strain was 0.75 AAs farther from the future than the vaccine strain.
Interestingly, the strains selected by the naive model were always better than the selected vaccine strain.
With an average improvement over the vaccine strains of 2.19 AAs, the naive model performed consistently better than the LBI-based model and nearly as well as the HI-based model.
These results were consistent with our earlier observations that the naive model often performs as well as biologically-informed models when estimating a single closest strain to the future.

>>>>

\begin{figure}
  \begin{center}
  \includegraphics[width=\textwidth]{figures/validation_figure_simulated-simulated_sample_3_test_tree-normalized_fitness.pdf}
  \caption{
  Test of best model for simulated populations (true fitness) using 9 years previously unobserved test data and fixed model coefficients.
  A) The correlation of log estimated clade frequency fold change, $\log_{10}{\frac{\hat{x}(t + \Delta{t})}{x(t)}}$, and log observed clade frequency fold change, $\log_{10}{\frac{x(t + \Delta{t})}{x(t)}}$, shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
  B) The rank of the estimated best strain based on its distance to the future in the best model was in the top 20th percentile for 89\% of 18 timepoints, confirming that the model makes a good choice when forced to select a single representative strain for the future population.
  C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (gray shading) based on 100 bootstraps.
  D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
  The corresponding results for the naive model are shown in Supplemental Fig.~\ref{sup_fig:test_of_naive_model_for_simulated_populations}.
  }
  \label{fig:test_of_best_model_for_simulated_populations}
  \end{center}
\end{figure}

@@

\begin{figure}
  \begin{center}
  \includegraphics[width=\textwidth]{figures/validation_figure_natural-natural_sample_1_with_90_vpm_sliding_test_tree-cTiter_x-ne_star.pdf}
  \caption{
  Test of best model for natural populations of H3N2 viruses, the composite model of HI antigenic novelty and mutational load.
  A) The correlation of estimated and observed clade frequency fold changes shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
  B) The rank of the estimated best strain based on its distance to the future for eight timepoints.
  The estimated best strain was in the top 20th percentile of observed closest strains for 100\% of timepoints.
  C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (gray shading) based on 100 bootstraps.
  D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
  The corresponding results for the naive model are shown in Supplemental Fig.~\ref{sup_fig:test_of_naive_model_for_natural_populations}.
  }
  \label{fig:test_of_best_model_for_natural_populations}
  \end{center}
\end{figure}

@@

\begin{figure}[htb]
  \includegraphics[width=\textwidth]{figures/vaccine-comparison.pdf}
  \caption{
    Observed distance to natural H3N2 populations one year into the future for each vaccine strain (green) and the observed (blue) and estimated closest strains to the future by the mutational load and LBI model (orange), the HI antigenic novelty and mutational load model (purple), and the naive model (black).
    Vaccine strains were assigned to the validation or test timepoint closest to the date they were selected by the WHO.
    The weighted distance to the future for each strain was calculated from their amino acid sequences and the frequencies and sequences of the corresponding population one year in the future.
    Vaccine strain names are abbreviated from A/Fujian/411/2002, A/Wellington/1/2004, A/California/7/2004, A/Wisconsin/67/2005, A/Brisbane/10/2007, A/Perth/16/2009, A/Victoria/361/2011, A/Texas/50/2012, A/Switzerland/9715293/2013, A/HongKong/4801/2014, A/Singapore/Infimh-16-0019/2016, and A/Switzerland/8060/2017.
  }
  \label{fig:vaccine_comparison}

  \figsupp[Relative improvement of model selections over vaccine strains.]{
    Relative distance to future H3N2 populations between vaccine strains and corresponding observed and estimated best strains at each timepoint as in Fig.~\ref{fig:vaccine_comparison}.
    Strains with relative distances greater than zero were farther from the future than the selected vaccine strain, while strains below zero were closer to the future.}{\includegraphics[width=\textwidth]{figures/vaccine-comparison-relative-distance.pdf}}\label{figsupp:vaccine_comparison_relative_distance}
\end{figure}

We further evaluated our models' ability to estimate the closest strain to the next season's H3N2 population by comparing our best models' selections to the WHO's vaccine strain selection.
For each season when the WHO selected a new vaccine strain and one year of future data existed in our validation or test periods, we measured the observed distance of that strain's sequence to the future and the corresponding distances to the future for the observed closest strains.
We compared these distances to those of the closest strains to the future as estimated by our best models for the validation period (mutational load and LBI) and the test period (HI antigenic novelty and mutational load).
The mutational load and LBI model selected strains that were as close or closer to the future than the corresponding vaccine strain for 10 (83\%) of the 12 seasons with vaccine updates (Fig.~\ref{fig:vaccine_comparison}).
On average, the strains selected by this model were closer to future than the vaccine strain by 1.93 AAs (Fig.~\ref{fig:vaccine_comparison}-Figure supplement~\ref{figsupp:vaccine_comparison_relative_distance}).
For the two seasons that the model selected more distant strains than the vaccine strain, the mean distance relative to the vaccine strain was 1.58 AAs.
The HI antigenic novelty and mutational load model performed similarly by identifying strains as close or closer to the future for 11 (92\%) seasons with an average improvement over the vaccine strains of 2.33 AAs.
For the one season that the model selected a more distant strain, that selected strain was 0.75 AAs farther from the future than the vaccine strain.
Interestingly, the strains selected by the naive model were always better than the selected vaccine strain.
With an average improvement over the vaccine strains of 2.19 AAs, the naive model performed consistently better than the LBI-based model and nearly as well as the HI-based model.
These results were consistent with our earlier observations that the naive model often performs as well as biologically-informed models when estimating a single closest strain to the future.
